\chapter*{Feature Engineering for Images (Part 2)}
\addcontentsline{toc}{chapter}{Feature Engineering for Images (Part 2)}

\section*{Image Binarization and Thresholding}

Many vision tasks require separating foreground from background before extracting higher-level structure. \textbf{Binarization} converts a grayscale image into a binary image by assigning each pixel to one of two classes.

\subsection*{Global Thresholding}

A single threshold $T$ is applied uniformly across the entire image:
$$
\text{Pixel}_{\text{out}} =
\begin{cases}
255 & \text{if } \text{Pixel}_{\text{in}} > T \\
0 & \text{otherwise}
\end{cases}
$$

\begin{lstlisting}[style=pythonstyle]
import cv2
import numpy as np

img = cv2.imread('page.png', cv2.IMREAD_GRAYSCALE)

# Global threshold at T=127: pixels > 127 become 255, else 0
ret, binary = cv2.threshold(img, 127, 255, cv2.THRESH_BINARY)
\end{lstlisting}

\begin{remark}
Global thresholding fails when the image has a lighting gradient (e.g., uneven shadows across a scanned document)---one region turns entirely white while the other turns entirely black.
\end{remark}

\subsection*{Adaptive (Local) Thresholding}

Instead of a single global $T$, a \textbf{local threshold} is computed independently for each pixel based on a neighborhood window (e.g., an $11 \times 11$ block):

\begin{description}
    \item[Mean Adaptive:] Threshold = mean intensity of the local neighborhood.
    \item[Gaussian Adaptive:] Threshold = Gaussian-weighted sum of neighborhood intensities, giving more influence to nearby pixels.
\end{description}

This naturally handles spatially varying illumination.

\begin{lstlisting}[style=pythonstyle]
# Adaptive mean thresholding (blockSize=11, C=2 subtracted from mean)
adapt_mean = cv2.adaptiveThreshold(
    img, 255, cv2.ADAPTIVE_THRESH_MEAN_C,
    cv2.THRESH_BINARY, blockSize=11, C=2
)

# Adaptive Gaussian thresholding
adapt_gauss = cv2.adaptiveThreshold(
    img, 255, cv2.ADAPTIVE_THRESH_GAUSSIAN_C,
    cv2.THRESH_BINARY, blockSize=11, C=2
)
\end{lstlisting}

\subsection*{Otsu's Binarization}

\textbf{Otsu's method} automatically discovers the optimal global threshold $T^*$ without manual tuning. It assumes the pixel histogram is \textbf{bimodal} (two peaks corresponding to foreground and background) and finds the $T$ that minimizes \textit{intra-class variance}---equivalently, maximizes \textit{inter-class variance}:
$$
T^* = \arg\min_T \;\bigl[\, \omega_0(T)\,\sigma_0^2(T) + \omega_1(T)\,\sigma_1^2(T) \,\bigr]
$$
where $\omega_k(T)$ and $\sigma_k^2(T)$ are the weight (pixel fraction) and variance of class $k$ at threshold $T$.

\begin{lstlisting}[style=pythonstyle]
# Otsu's method: pass 0 as threshold, OpenCV finds optimal T*
ret, otsu = cv2.threshold(
    img, 0, 255, cv2.THRESH_BINARY + cv2.THRESH_OTSU
)
print(f"Otsu's optimal threshold: {ret}")  # e.g., 142.0
\end{lstlisting}

\section*{Corner Detection}

To track objects across frames or match features between images, we need to identify points that are \textbf{distinct} and \textbf{repeatable}. Not all image regions are equally useful:

\begin{description}
    \item[Flat Regions:] No intensity gradient in any direction. Useless for tracking---shifting a window produces no measurable change.
    \item[Edges:] Strong gradient in one direction only. Subject to the \textbf{aperture problem}: we cannot determine position along the edge.
    \item[Corners:] High gradients in \textit{multiple} directions. Highly distinct and ideal as keypoints.
\end{description}

\subsection*{Harris Corner Detector}

The Harris detector identifies corners by analyzing how the \textbf{sum of squared differences (SSD)} changes when a local window is shifted by $(u, v)$:
$$
E(u,v) = \sum_{(x,y) \in W} \bigl[ I(x+u, y+v) - I(x,y) \bigr]^2
$$

A first-order Taylor expansion yields the approximation $E(u,v) \approx \begin{bmatrix} u & v \end{bmatrix} M \begin{bmatrix} u \\ v \end{bmatrix}$, where $M$ is the \textbf{Structure Tensor} (Second Moment Matrix) computed from image derivatives $I_x, I_y$ over window $W$:
$$
M = \sum_{(x,y) \in W} \begin{bmatrix} I_x^2 & I_x I_y \\ I_x I_y & I_y^2 \end{bmatrix}
$$

The eigenvalues $\lambda_1, \lambda_2$ of $M$ encode the local gradient structure:

\begin{description}
    \item[Corner ($\lambda_1$ large, $\lambda_2$ large):] Intensity changes significantly in all shift directions.
    \item[Edge ($\lambda_1 \gg \lambda_2$ or vice versa):] Change in one direction only.
    \item[Flat ($\lambda_1 \approx \lambda_2 \approx 0$):] No significant change.
\end{description}

\begin{remark}
Computing eigenvalues directly is expensive. Harris instead uses the \textbf{corner response function}:
$$R = \det(M) - k\,(\text{trace}(M))^2 = \lambda_1 \lambda_2 - k(\lambda_1 + \lambda_2)^2$$
where $k \in [0.04, 0.06]$. A large positive $R$ indicates a corner; $R < 0$ indicates an edge; $|R|$ small indicates a flat region.
\end{remark}

\begin{lstlisting}[style=pythonstyle]
import cv2
import numpy as np

img = cv2.imread('chessboard.png')
gray = cv2.cvtColor(img, cv2.COLOR_BGR2GRAY).astype(np.float32)

# Harris corner detection (blockSize=2, ksize=3, k=0.04)
harris = cv2.cornerHarris(gray, blockSize=2, ksize=3, k=0.04)

# Dilate to enlarge corner markers for visibility
harris = cv2.dilate(harris, None)

# Mark corners in red where response > 1% of max
img[harris > 0.01 * harris.max()] = [0, 0, 255]
\end{lstlisting}

\section*{The Feature Matching Pipeline}

Given keypoints in two images, the classical pipeline for establishing correspondences (e.g., for panorama stitching, 3D reconstruction, or object recognition) consists of three stages:

\begin{enumerate}
    \item \textbf{Detection:} Identify repeatable interest points (corners, blobs) in each image.
    \item \textbf{Description:} For each keypoint, compute a descriptor vector encoding the local neighborhood. The descriptor must be robust to rotation, illumination changes, and ideally scale.
    \item \textbf{Matching:} Compare descriptors between images using a distance metric (Euclidean for continuous descriptors, Hamming for binary) and select the closest pairs as correspondences.
\end{enumerate}

\section*{Advanced Feature Descriptors}

Harris corners are rotation-invariant but \textbf{not scale-invariant}: zoom into a corner far enough and it becomes an edge. Modern classical descriptors address this limitation.

\subsection*{SIFT (Scale-Invariant Feature Transform)}

SIFT achieves invariance to scale, rotation, and moderate affine distortion through a multi-stage pipeline:

\begin{description}
    \item[Detection (DoG):] Constructs a \textbf{scale space} by convolving the image with Gaussians at increasing $\sigma$. The \textbf{Difference of Gaussians (DoG)} $D(x,y,\sigma) = G(x,y,k\sigma) - G(x,y,\sigma)$ approximates the scale-normalized Laplacian. Keypoints are local extrema in $(x, y, \sigma)$ space---points that persist across scales.
    \item[Orientation Assignment:] For each keypoint, a histogram of gradient orientations in its neighborhood determines the \textbf{dominant orientation}. All subsequent computations are relative to this orientation, ensuring rotation invariance.
    \item[Descriptor:] A $16 \times 16$ patch around the keypoint is divided into a $4 \times 4$ grid of cells. Each cell contributes an 8-bin \textbf{Histogram of Oriented Gradients (HOG)}. Concatenation yields a $4 \times 4 \times 8 = \mathbf{128}$-dimensional continuous feature vector, normalized to unit length for illumination invariance.
\end{description}

\begin{lstlisting}[style=pythonstyle]
import cv2

img = cv2.imread('building.jpg')
gray = cv2.cvtColor(img, cv2.COLOR_BGR2GRAY)

# Create SIFT detector and compute keypoints + descriptors
sift = cv2.SIFT_create()
keypoints, descriptors = sift.detectAndCompute(gray, None)
# descriptors.shape = (N, 128) -- N keypoints, 128-dim each

# Draw keypoints with orientation and scale
output = cv2.drawKeypoints(
    gray, keypoints, img,
    flags=cv2.DRAW_MATCHES_FLAGS_DRAW_RICH_KEYPOINTS
)
\end{lstlisting}

\subsection*{ORB (Oriented FAST and Rotated BRIEF)}

ORB was designed as an efficient, royalty-free alternative to SIFT, widely used in real-time robotics and mobile applications.

\begin{description}
    \item[FAST Detection:] Identifies keypoints by comparing a center pixel's intensity to a ring of 16 surrounding pixels. If $N$ contiguous pixels on the ring are all significantly brighter (or all darker) than the center, it is classified as a corner. FAST is extremely efficient but lacks orientation information.
    \item[Orientation (``Oriented''):] ORB adds rotation invariance by computing the \textbf{intensity centroid} of the patch and defining orientation as the direction from the corner to the centroid.
    \item[BRIEF Description:] Generates a \textbf{binary descriptor} (256 bits by default) by performing pairwise intensity comparisons at pre-selected pixel locations in the neighborhood. Each comparison yields one bit: 1 if $I(p_1) > I(p_2)$, else 0.
    \item[Matching with Hamming Distance:] Because descriptors are binary strings, matching reduces to XOR followed by popcount---executed in a single CPU instruction. This is orders of magnitude faster than Euclidean distance on 128-dim floats.
\end{description}

\begin{lstlisting}[style=pythonstyle]
import cv2

img = cv2.imread('building.jpg', cv2.IMREAD_GRAYSCALE)

# Create ORB detector (default: 500 keypoints, 256-bit descriptors)
orb = cv2.ORB_create(nfeatures=1000)
keypoints, descriptors = orb.detectAndCompute(img, None)
# descriptors.shape = (N, 32) -- 32 bytes = 256 bits per keypoint
\end{lstlisting}

\subsection*{Feature Matching in Practice}

Given descriptors from two images, a \textbf{Brute-Force Matcher} compares every descriptor in image A against every descriptor in image B:

\begin{lstlisting}[style=pythonstyle]
import cv2

img1 = cv2.imread('scene1.jpg', cv2.IMREAD_GRAYSCALE)
img2 = cv2.imread('scene2.jpg', cv2.IMREAD_GRAYSCALE)

# --- SIFT matching (L2 norm on 128-dim float vectors) ---
sift = cv2.SIFT_create()
kp1, des1 = sift.detectAndCompute(img1, None)
kp2, des2 = sift.detectAndCompute(img2, None)

bf = cv2.BFMatcher(cv2.NORM_L2, crossCheck=True)  # L2 for SIFT
matches = bf.match(des1, des2)
matches = sorted(matches, key=lambda m: m.distance)  # best first

# --- ORB matching (Hamming distance on binary descriptors) ---
orb = cv2.ORB_create(nfeatures=1000)
kp1, des1 = orb.detectAndCompute(img1, None)
kp2, des2 = orb.detectAndCompute(img2, None)

bf = cv2.BFMatcher(cv2.NORM_HAMMING, crossCheck=True)  # Hamming for ORB
matches = bf.match(des1, des2)
matches = sorted(matches, key=lambda m: m.distance)
\end{lstlisting}

\begin{remark}[SIFT vs.\ ORB Trade-off]
SIFT produces highly discriminative 128-dim float descriptors matched with $L_2$ distance---accurate but slow ($O(128n^2)$ for brute-force). ORB produces compact 256-bit binary descriptors matched with Hamming distance via hardware XOR---roughly 10--100$\times$ faster. For real-time systems (SLAM, AR), ORB is preferred; for offline tasks requiring maximum accuracy (3D reconstruction, archival matching), SIFT remains the gold standard among classical methods.
\end{remark}
