\chapter*{Logistic Regression}
\addcontentsline{toc}{chapter}{Logistic Regression}

\section*{A Probabilistic View of Classification}

Previous chapters framed classification geometrically: find a hyperplane that separates classes. Logistic Regression introduces a \textbf{probabilistic formulation}---instead of predicting a hard label $y \in \{0,1\}$, we model the \textbf{conditional probability} that the label is $y$ given input $\xx$:
$$P(y \mid \xx)$$
This is strictly more informative: it yields the most likely label \textit{and} quantifies the model's uncertainty in that prediction.

\begin{description}
    \item[Generative Models:] Learn the joint $P(\xx, y)$ or the class-conditional $P(\xx \mid y)$, then apply Bayes' rule to obtain $P(y \mid \xx) = \frac{P(\xx \mid y)\,P(y)}{P(\xx)}$. They model \textit{how the data was generated} (e.g., Naive Bayes).
    \item[Discriminative Models:] Learn $P(y \mid \xx)$ directly, bypassing the full data distribution. \textbf{Logistic Regression is a discriminative model.}
\end{description}

\section*{The Logistic Sigmoid Function}

A linear model computes a score $z = \ww^T\xx + w_0 \in (-\infty, +\infty)$. To interpret this as a probability, we need a mapping to $[0,1]$.

\begin{definition}[Logistic Sigmoid Function]
The sigmoid function $\sigma(z)$ squashes any real number into the interval $(0,1)$:
$$\sigma(z) = \frac{1}{1 + e^{-z}} = \frac{e^z}{1 + e^z}$$
\end{definition}

\textbf{Key Properties of $\sigma(z)$:}
\begin{itemize}
    \item $\sigma(z) \to 1$ as $z \to +\infty$; \quad $\sigma(z) \to 0$ as $z \to -\infty$; \quad $\sigma(0) = 0.5$.
    \item \textbf{Symmetry:} $\sigma(-z) = 1 - \sigma(z)$.
    \item Nearly linear around $z=0$, so small changes in the score produce proportional changes in probability.
    \item Flattens toward the extremes, squashing outlier scores toward 0 or 1.
    \item \textbf{Differentiable}, with a clean derivative: $\sigma'(z) = \sigma(z)\,(1 - \sigma(z))$.
\end{itemize}

\begin{lstlisting}[style=pythonstyle]
import numpy as np
import matplotlib.pyplot as plt

# Plot the sigmoid function
z = np.linspace(-7, 7, 300)
sigma = 1 / (1 + np.exp(-z))          # sigmoid formula

plt.plot(z, sigma, linewidth=2)
plt.axhline(0.5, color='gray', linestyle='--', linewidth=0.8)
plt.axvline(0.0, color='gray', linestyle='--', linewidth=0.8)
plt.xlabel("z (logit)")
plt.ylabel(r"$\sigma(z)$")
plt.title("Logistic Sigmoid Function")
plt.show()
\end{lstlisting}

\begin{remark}[Logit: The Inverse of the Sigmoid]
The input $z$ to the sigmoid is called the \textbf{logit}. The logit function is the \textit{inverse} of the sigmoid and equals the log of the odds ratio:
$$\text{logit}(p) = \sigma^{-1}(p) = \ln\frac{p}{1-p}$$
Interpreting $z$ as a log-odds gives probabilistic meaning to the linear score: a score of $z=2$ means the log-odds of class 1 are $2$, i.e., $P(y{=}1)/P(y{=}0) = e^2 \approx 7.4$.
\end{remark}

\section*{The Logistic Regression Model}

For binary classification ($y \in \{0, 1\}$), Logistic Regression models the probability of the positive class as:
$$P(y{=}1 \mid \xx) = \sigma(\ww^T\xx + w_0) = \frac{1}{1 + e^{-(\ww^T\xx + w_0)}}$$
and consequently:
$$P(y{=}0 \mid \xx) = 1 - \sigma(\ww^T\xx + w_0) = \sigma(-(\ww^T\xx + w_0))$$
Both cases can be written compactly using the Bernoulli form:
$$P(y \mid \xx, \ww, w_0) = \hat{y}^{\,y}\,(1-\hat{y})^{1-y}, \quad \hat{y} = \sigma(\ww^T\xx + w_0)$$

\begin{remark}[The Decision Boundary Is Linear]
To make a hard prediction, we apply a threshold at $0.5$: predict $y{=}1$ if $P(y{=}1 \mid \xx) \geq 0.5$. Since $\sigma(z) \geq 0.5 \iff z \geq 0$, the rule simplifies to:
$$\text{Predict } 1 \;\text{ if }\; \ww^T\xx + w_0 \geq 0$$
Despite the non-linear sigmoid, \textbf{the decision boundary is the hyperplane} $\ww^T\xx + w_0 = 0$---exactly where both classes are equiprobable.
\end{remark}

\begin{lstlisting}[style=pythonstyle]
from sklearn.linear_model import LogisticRegression
from sklearn.datasets import make_classification
from sklearn.model_selection import train_test_split

# Generate a binary classification dataset
X, y = make_classification(n_samples=300, n_features=2,
                           n_redundant=0, random_state=42)
X_train, X_test, y_train, y_test = train_test_split(
    X, y, test_size=0.3, random_state=42)

# Fit logistic regression
model = LogisticRegression()
model.fit(X_train, y_train)

print(f"Weights w:  {model.coef_[0]}")
print(f"Bias w_0:   {model.intercept_[0]:.4f}")
print(f"Accuracy:   {model.score(X_test, y_test):.3f}")
\end{lstlisting}

\begin{lstlisting}[style=pythonstyle]
# predict_proba returns [P(y=0|x), P(y=1|x)] for each sample
probs = model.predict_proba(X_test[:5])
for i, (p0, p1) in enumerate(probs):
    label = model.predict(X_test[i:i+1])[0]
    print(f"Sample {i}: P(0)={p0:.3f}  P(1)={p1:.3f} -> y_hat={label}")
\end{lstlisting}

\section*{Training: Maximum Conditional Likelihood}

Because Logistic Regression outputs probabilities, we train it by \textbf{Maximum Conditional Likelihood Estimation (MCLE)}: find $\ww, w_0$ that make the observed labels most probable under the model.

Given $N$ i.i.d.\ training pairs, the conditional likelihood is:
$$L(\ww, w_0) = \prod_{i=1}^{N} P(y^{(i)} \mid \xx^{(i)}, \ww, w_0) = \prod_{i=1}^{N} \hat{y}^{(i)^{\,y^{(i)}}} (1-\hat{y}^{(i)})^{1-y^{(i)}}$$
Taking the negative log converts the product to a sum and flips to minimization.

\begin{definition}[Binary Cross-Entropy / Log Loss]
The Logistic Regression cost function is the negative log-likelihood, averaged over $N$ samples:
$$\mathcal{L}(\ww, w_0) = -\frac{1}{N}\sum_{i=1}^{N}\Big[y^{(i)}\log\hat{y}^{(i)} + (1-y^{(i)})\log(1-\hat{y}^{(i)})\Big]$$
where $\hat{y}^{(i)} = \sigma(\ww^T\xx^{(i)} + w_0)$.
\end{definition}

\textbf{Why this loss works:} When $y=1$, only the $-\log\hat{y}$ term is active---if the model confidently predicts $\hat{y} \approx 0$, the loss $-\log(\hat{y}) \to \infty$, heavily penalizing the wrong confident prediction. Symmetrically for $y=0$.

\begin{remark}[Equivalence to ERM]
Minimizing the negative log-likelihood is equivalent to Empirical Risk Minimization (as established in the ERM framework) with the log-loss as the per-sample loss function. The probabilistic and optimization views of ML converge here.
\end{remark}

\section*{Optimization}

Setting $\nabla \mathcal{L} = 0$ yields no closed-form solution due to the non-linear $e^{-z}$ term inside $\sigma$. We must use \textbf{iterative optimization}.

\textbf{Key fact:} The log-loss is a \textbf{strictly convex} function of $\ww$ (it is a linear combination of convex functions). This guarantees:
\begin{itemize}
    \item Every local minimum is the global minimum.
    \item Gradient descent converges to the unique optimal $\ww^*$ (given appropriate learning rate).
\end{itemize}

The gradient of the log-likelihood (for gradient \textit{ascent}) has a remarkably clean form:
$$\nabla_{\ww}\,\ell(\ww) = \sum_{i=1}^{N}\big(y^{(i)} - \sigma(\ww^T\xx^{(i)} + w_0)\big)\,\xx^{(i)}$$
Each term is the prediction error $(y - \hat{y})$ scaled by the input $\xx$---the same structure as in linear regression.

\section*{Regularization}

\begin{remark}[Why Regularize?]
If training data is \textbf{perfectly linearly separable}, the model can drive the loss to zero by pushing $\hat{y} \to 1$ or $\hat{y} \to 0$ for every sample. This requires $\|\ww\| \to \infty$, producing an infinitely steep sigmoid that memorizes the training data but generalizes poorly.
\end{remark}

Adding a penalty term prevents weight explosion:
\begin{description}
    \item[$L_2$ (Ridge):] $\mathcal{L}_{\text{reg}} = \mathcal{L}(\ww) + \beta\sum_j w_j^2$ \quad---\quad prefers many small weights; smooth optimization.
    \item[$L_1$ (Lasso):] $\mathcal{L}_{\text{reg}} = \mathcal{L}(\ww) + \alpha\sum_j |w_j|$ \quad---\quad drives some weights to exactly zero (feature selection).
\end{description}

In \texttt{sklearn}, regularization strength is controlled by the parameter \texttt{C}, where $\texttt{C} = 1/\lambda$. Smaller \texttt{C} means stronger regularization.

\begin{lstlisting}[style=pythonstyle]
from sklearn.linear_model import LogisticRegression
import numpy as np

# Compare regularization strengths via the C parameter
# C = 1/lambda: smaller C = stronger regularization
for C_val in [0.01, 1.0, 100.0]:
    model = LogisticRegression(C=C_val, penalty='l2', max_iter=1000)
    model.fit(X_train, y_train)
    w_norm = np.linalg.norm(model.coef_)
    acc = model.score(X_test, y_test)
    print(f"C={C_val:6.2f}  ||w||={w_norm:.4f}  Acc={acc:.3f}")
# Strong regularization (small C) -> smaller weights, may underfit
# Weak regularization (large C) -> larger weights, may overfit
\end{lstlisting}

\section*{Multi-Class Extension: Softmax Regression}

Binary Logistic Regression generalizes to $K$ classes via the \textbf{softmax function}. Instead of a single weight vector, we learn a weight matrix $\mathbf{W} = [\ww_1, \dots, \ww_K]$ with one vector per class:
$$P(y{=}k \mid \xx, \mathbf{W}) = \text{softmax}(\mathbf{W}, \xx, k) = \frac{e^{\ww_k^T\xx}}{\sum_{j=1}^{K} e^{\ww_j^T\xx}}$$
The softmax normalizes the $K$ raw scores into a valid probability distribution that sums to 1. For $K{=}2$, this reduces exactly to the sigmoid.

The corresponding loss is the \textbf{categorical cross-entropy}:
$$\mathcal{L} = -\frac{1}{N}\sum_{i=1}^{N}\sum_{k=1}^{K}\mathbf{1}[y^{(i)}{=}k]\;\log P(y{=}k \mid \xx^{(i)})$$

\begin{lstlisting}[style=pythonstyle]
from sklearn.datasets import load_iris
from sklearn.linear_model import LogisticRegression
from sklearn.model_selection import train_test_split

# Iris: 3 classes -> softmax is used automatically
X, y = load_iris(return_X_y=True)
X_tr, X_te, y_tr, y_te = train_test_split(X, y, test_size=0.3,
                                           random_state=0)

# multi_class='multinomial' uses softmax (default in sklearn)
model = LogisticRegression(multi_class='multinomial', max_iter=200)
model.fit(X_tr, y_tr)

print(f"Classes:  {model.classes_}")
print(f"Accuracy: {model.score(X_te, y_te):.3f}")
# Each row of predict_proba sums to 1 across all K classes
print(f"Probs (first sample): {model.predict_proba(X_te[:1])[0]}")
\end{lstlisting}
