\chapter*{Support Vector Machines (SVM)}
\addcontentsline{toc}{chapter}{Support Vector Machines (SVM)}

\section*{The Optimal Decision Boundary}

For linearly separable data, infinitely many hyperplanes can perfectly separate two classes. \textbf{Support Vector Machines (SVMs)} answer the question: \textit{which of these boundaries generalizes best?}

The SVM principle is intuitive: the best decision boundary is the one \textbf{furthest from the closest training points} of any class. A classifier with a wider ``safe zone'' around its boundary is less likely to misclassify unseen data that falls near the border. This safe zone is formalized as the \textbf{geometric margin}.

\begin{definition}[Geometric Margin]
The \textbf{margin} $\gamma$ is the perpendicular distance between the decision hyperplane $\mathbf{w}^T\xx + b = 0$ and the closest data points from each class. The total margin width is $\gamma = \frac{2}{\|\mathbf{w}\|}$. SVM seeks the hyperplane that \textbf{maximizes} $\gamma$.
\end{definition}

\begin{definition}[Support Vectors]
The \textbf{support vectors} are the data points lying exactly on the margin boundaries (i.e., satisfying $y^{(i)}(\mathbf{w}^T\xx^{(i)} + b) = 1$). The optimal hyperplane depends \textit{only} on these critical points---removing or moving any non-support-vector training point leaves the decision boundary unchanged.
\end{definition}

\begin{lstlisting}[style=pythonstyle]
# Train a linear SVM classifier
from sklearn.svm import SVC
from sklearn.model_selection import train_test_split
from sklearn.metrics import accuracy_score

X_train, X_test, y_train, y_test = train_test_split(
    X, y, test_size=0.25, random_state=42)

svm_clf = SVC(kernel='linear')       # Linear (hard/soft) SVM
svm_clf.fit(X_train, y_train)
y_pred = svm_clf.predict(X_test)
print(f"Accuracy: {accuracy_score(y_test, y_pred):.3f}")
\end{lstlisting}

\begin{lstlisting}[style=pythonstyle]
# Inspect support vectors after fitting
print(f"Support vectors per class: {svm_clf.n_support_}")
print(f"Total support vectors: {len(svm_clf.support_vectors_)}")

# Access the learned hyperplane: w^T x + b = 0
print(f"Weights w: {svm_clf.coef_[0]}")
print(f"Intercept b: {svm_clf.intercept_[0]:.4f}")
\end{lstlisting}

\section*{Hard Margin SVM}

\textbf{Hard Margin SVM} enforces that every training point is correctly classified and lies outside the margin band. Since the margin width is $\frac{2}{\|\mathbf{w}\|}$, maximizing the margin is equivalent to minimizing $\|\mathbf{w}\|^2$. The resulting \textbf{quadratic program} (convex, with linear inequality constraints) is:

$$\min_{\mathbf{w}, b} \;\frac{1}{2} \|\mathbf{w}\|^2 \qquad \text{s.t.} \quad y^{(i)}(\mathbf{w}^T\xx^{(i)} + b) \geq 1, \quad i = 1, \ldots, m$$

\begin{remark}[Why $\frac{1}{2}\|\mathbf{w}\|^2$?]
The geometric margin $\gamma = \frac{2c}{\|\mathbf{w}\|}$ is scaling-invariant in $\mathbf{w}$, so we can fix $c = 1$ without loss of generality. Squaring removes the square root (a monotonic transformation that preserves the optimum) and the $\frac{1}{2}$ simplifies the gradient to $\mathbf{w}$.
\end{remark}

\begin{remark}
Hard Margin SVM requires \textbf{perfect linear separability}. A single outlier crossing the boundary can make the problem infeasible or drastically skew the solution.
\end{remark}

\subsection*{Primal and Dual Formulations}

The \textbf{primal} solves for the $d$-dimensional weight vector $\mathbf{w}$. Introducing Lagrange multipliers $\alpha_i \geq 0$ for each constraint yields the equivalent \textbf{dual} problem over $m$ variables:

$$\max_{\boldsymbol{\alpha}} \;\sum_{i=1}^{m} \alpha_i - \frac{1}{2}\sum_{i,j=1}^{m} \alpha_i \alpha_j \, y^{(i)} y^{(j)} \, (\xx^{(i)})^T\xx^{(j)} \qquad \text{s.t.} \quad \sum_{i=1}^{m} \alpha_i y^{(i)} = 0, \;\; \alpha_i \geq 0$$

At the solution, $\mathbf{w} = \sum_{i=1}^{m} \alpha_i y^{(i)} \xx^{(i)}$. Only support vectors have $\alpha_i > 0$; all other training points have $\alpha_i = 0$ and are irrelevant. The bias is recovered as $b = y^{(k)} - \mathbf{w}^T\xx^{(k)}$ for any support vector $k$.

\section*{Soft Margin SVM}

When data is not perfectly separable (or contains noisy outliers), we relax the hard constraints by introducing \textbf{slack variables} $\xi_i \geq 0$ that allow individual points to violate the margin or be misclassified:

$$\min_{\mathbf{w}, b, \boldsymbol{\xi}} \;\frac{1}{2}\|\mathbf{w}\|^2 + C \sum_{i=1}^{m} \xi_i \qquad \text{s.t.} \quad y^{(i)}(\mathbf{w}^T\xx^{(i)} + b) \geq 1 - \xi_i, \;\; \xi_i \geq 0$$

The slack variable $\xi_i$ has a direct geometric meaning:
\begin{itemize}
    \item $\xi_i = 0$: point $\xx^{(i)}$ is correctly classified and outside the margin.
    \item $0 < \xi_i < 1$: correctly classified but inside the margin (margin violation).
    \item $\xi_i \geq 1$: misclassified.
\end{itemize}

\begin{description}
    \item[The $C$ Parameter:] $C$ controls the trade-off between margin width and constraint violations.
    \begin{itemize}
        \item \textbf{Large $C$:} High penalty for violations $\Rightarrow$ narrower margin, fewer errors on training data (risk of overfitting).
        \item \textbf{Small $C$:} Low penalty for violations $\Rightarrow$ wider margin, more tolerance for misclassifications (better generalization, risk of underfitting).
    \end{itemize}
    As $C \to \infty$, the soft-margin formulation recovers the hard-margin SVM.
\end{description}

The soft-margin dual differs from the hard-margin dual only by an upper bound on each multiplier: $0 \leq \alpha_i \leq C$.

\begin{lstlisting}[style=pythonstyle]
# Effect of C: small C = wide margin, large C = narrow margin
from sklearn.svm import SVC

svm_small_C = SVC(kernel='linear', C=0.01)  # Wide margin
svm_large_C = SVC(kernel='linear', C=100)   # Narrow margin

svm_small_C.fit(X_train, y_train)
svm_large_C.fit(X_train, y_train)

# Small C: more support vectors, wider margin
print(f"C=0.01 SVs: {sum(svm_small_C.n_support_)}")
# Large C: fewer support vectors, tighter fit
print(f"C=100  SVs: {sum(svm_large_C.n_support_)}")
\end{lstlisting}

\section*{SVM and the Hinge Loss}

Soft Margin SVM can be understood through the \textbf{Empirical Risk Minimization (ERM)} framework. Training a soft-margin SVM is mathematically equivalent to minimizing the \textbf{hinge loss} with $L_2$ regularization:

$$\min_{\mathbf{w}} \;\sum_{i=1}^{m} \max\!\bigl(0,\; 1 - y^{(i)}(\mathbf{w}^T\xx^{(i)} + b)\bigr) + \lambda \|\mathbf{w}\|^2$$

The equivalence follows by identifying the slack variables as $\xi_i = \max(0, 1 - y^{(i)} f_{\mathbf{w}}(\xx^{(i)}))$ and the regularization strength as $\lambda = \frac{1}{2C}$.

\begin{remark}[Hinge Loss vs.\ 0--1 Loss]
The hinge loss is a convex upper bound on the 0--1 loss. It penalizes not only misclassifications but also correct predictions that fall inside the margin ($y^{(i)} f_{\mathbf{w}}(\xx^{(i)}) < 1$), pushing the algorithm to achieve a margin of at least 1 on every point. Keeping $\|\mathbf{w}\|$ small (via regularization) translates directly to a larger geometric margin, since $\gamma = \frac{2}{\|\mathbf{w}\|}$.
\end{remark}

\begin{remark}[SVM vs.\ Logistic Regression]
Both are linear classifiers, but they differ in their loss functions. Logistic regression uses the \textbf{log loss} (cross-entropy), which always imposes a penalty---even for correctly classified points far from the boundary. The hinge loss imposes \textbf{zero penalty} once a point is outside the margin ($y \cdot f(\xx) \geq 1$), making SVMs sparser: only support vectors influence the solution.
\end{remark}

\section*{The Kernel Trick}

When data is highly non-linear, no hyperplane in the original feature space can achieve good separation. We could map inputs to a higher-dimensional space $\phi(\xx)$ where the data becomes separable, but explicitly computing $\phi(\xx)$ is often prohibitively expensive.

\begin{definition}[The Kernel Trick]
In the dual formulation, training data appears only through dot products $(\xx^{(i)})^T\xx^{(j)}$. A \textbf{kernel function} $K(\xx^{(i)}, \xx^{(j)}) = \phi(\xx^{(i)})^T \phi(\xx^{(j)})$ computes the inner product in the transformed space \textit{without ever explicitly computing $\phi$}. Replacing dot products with $K$ in the dual lets SVM learn non-linear boundaries at the cost of a linear kernel evaluation.
\end{definition}

The prediction for a new input $\mathbf{z}$ becomes:
$$y(\mathbf{z}) = \text{sign}\!\left(b + \sum_{i \in SV} \alpha_i \, y^{(i)} \, K(\xx^{(i)}, \mathbf{z})\right)$$

\textbf{Common kernel functions:}
\begin{itemize}
    \item \textbf{Linear:} $K(\xx_i, \xx_j) = \xx_i^T \xx_j$ \;---\; no transformation (standard SVM).
    \item \textbf{Polynomial:} $K(\xx_i, \xx_j) = (\xx_i^T \xx_j + c)^d$ \;---\; maps to polynomial feature space of degree $d$.
    \item \textbf{RBF / Gaussian:} $K(\xx_i, \xx_j) = \exp\!\bigl(-\gamma \|\xx_i - \xx_j\|^2\bigr)$ \;---\; maps to an \textit{infinite}-dimensional space. The parameter $\gamma$ controls how quickly similarity decays with distance.
\end{itemize}

\begin{lstlisting}[style=pythonstyle]
# Kernel SVM with RBF (Gaussian) kernel
from sklearn.svm import SVC
from sklearn.preprocessing import StandardScaler
from sklearn.pipeline import make_pipeline

# RBF kernel is sensitive to feature scale -- always standardize
rbf_svm = make_pipeline(
    StandardScaler(),
    SVC(kernel='rbf', C=1.0, gamma='scale')  # RBF kernel
)
rbf_svm.fit(X_train, y_train)
print(f"RBF SVM accuracy: {rbf_svm.score(X_test, y_test):.3f}")

# Polynomial kernel of degree 3
poly_svm = make_pipeline(
    StandardScaler(),
    SVC(kernel='poly', degree=3, C=1.0)
)
poly_svm.fit(X_train, y_train)
print(f"Poly SVM accuracy: {poly_svm.score(X_test, y_test):.3f}")
\end{lstlisting}

\section*{Support Vector Regression (SVR)}

SVM concepts extend naturally from classification to regression. In \textbf{Support Vector Regression (SVR)}, the margin concept is replaced by an \textbf{$\varepsilon$-insensitive tube} surrounding the regression function $f(\xx)$.

\begin{description}
    \item[$\varepsilon$-Insensitive Loss:] Predictions falling \textit{inside} the tube (within $\varepsilon$ of the true target) incur \textbf{zero loss}. Only predictions \textit{outside} the tube pay a linear penalty via slack variables $\xi_i, \hat{\xi}_i \geq 0$ (one for each direction---above and below the tube).
\end{description}

The SVR primal optimization problem is:
$$\min_{\mathbf{w}, b, \boldsymbol{\xi}, \hat{\boldsymbol{\xi}}} \;\frac{1}{2}\|\mathbf{w}\|^2 + C \sum_{i=1}^{m} (\xi_i + \hat{\xi}_i)$$
$$\text{s.t.} \quad y^{(i)} - \mathbf{w}^T\xx^{(i)} - b \leq \varepsilon + \xi_i, \quad \mathbf{w}^T\xx^{(i)} + b - y^{(i)} \leq \varepsilon + \hat{\xi}_i, \quad \xi_i, \hat{\xi}_i \geq 0$$

Like classification SVMs, SVR relies on support vectors (the points on or outside the tube boundary) and can be fully \textbf{kernelized} to fit non-linear regression curves.

\begin{lstlisting}[style=pythonstyle]
# Support Vector Regression (SVR)
from sklearn.svm import SVR
from sklearn.preprocessing import StandardScaler
from sklearn.pipeline import make_pipeline

svr_model = make_pipeline(
    StandardScaler(),
    SVR(kernel='rbf', C=100, epsilon=0.1)  # eps-insensitive tube
)
svr_model.fit(X_train, y_train)
y_pred = svr_model.predict(X_test)

from sklearn.metrics import mean_squared_error
print(f"SVR MSE: {mean_squared_error(y_test, y_pred):.4f}")
\end{lstlisting}
